---
pagetitle: 'Svensk Fågeltaxering'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```


## Introduction

This exercise is designed to familiarize you with the data from *The Swedish Bird Inventory (Svensk Fågeltaxering)*. On [this](https://www.fageltaxering.lu.se/) website, you can read more about the dataset (field protocol, history, etc.). A secondary aim is to continue improving your skills with R.

## Getting started
### Working directory, Project folders, and Reading data

* Remember that your working directory is usually the folder where your R script or R project is kept.
* If you place the data for today's exercise in the same folder as the project you created for the exercises in this class, you can simply continue working in that project and the data should be available.
* Otherwise refer to the "Introduction to R" exercise for more details on working directories and loading data in R.

### Download bird data for this exercise and read into R

You can download the data for this exercise [here](https://clausrueffler.github.io/svensk-fageltaxering/downloads/bird_data_subset.csv). Alternatively, you can download the data as an R-data file [here](https://clausrueffler.github.io/svensk-fageltaxering/downloads/bird_data_subset.RDA). 

The dataset that you can download here is a reduced version of the full dataset, which is available [from the GBIF website](https://www.gbif.org/dataset/91fa1a0d-a208-40aa-8a6e-f2c0beb9b253).

Move the downloaded *.csv* file into your project folder (working directory) and then read the data into R:

```{r, echo=T, eval=T}
birds <- read.csv("bird_data_subset.csv", stringsAsFactors = FALSE)

# or
birds <- readRDS("bird_data_subset.RDA")
```

```{r, echo=T, eval=T}

```

## Trends of farmland and forest bird species
### Background

Below, you can find two lists of ten common bird species associated with farmlands and forest habitats, respectively. The dataset we are working with here has been reduced to include only these species and only a few columns with the most relevant information. *Click on the names to get more information on the species and to see some nice pictures!*

#### Farmland birds:
* [Common Kestrel (*Falco tinnunculus*)](https://en.wikipedia.org/wiki/Common_kestrel)
* [Northern Lapwing (*Vanellus vanellus*)](https://en.wikipedia.org/wiki/Northern_lapwing)
* [Skylark (*Alauda arvensis*)](https://en.wikipedia.org/wiki/Eurasian_skylark)
* [Whinchat (*Saxicola rubetra*)](https://en.wikipedia.org/wiki/Whinchat)
* [Meadow pipit (*Anthus pratensis*)](https://en.wikipedia.org/wiki/Meadow_pipit)
* [Yellow wagtail (*Motacilla flava*)](https://en.wikipedia.org/wiki/Western_yellow_wagtail)
* [Yellowhammer (*Emberiza citrinella*)](https://en.wikipedia.org/wiki/Yellowhammer)
* [Tree sparrow (*Passer montanus*)](https://en.wikipedia.org/wiki/Eurasian_tree_sparrow)
* [Starling (*Sturnus vulgaris*)](https://en.wikipedia.org/wiki/Common_starling)
* [Common whitethroat (*Sylvia communis*)](https://en.wikipedia.org/wiki/Common_whitethroat)

#### Forest birds:
* [Eurasian sparrowhawk (*Accipiter nisus*)](https://en.wikipedia.org/wiki/Eurasian_sparrowhawk)
* [Euroasian jay (*Garrulus glandarius*)](https://en.wikipedia.org/wiki/Eurasian_jay)
* [Greater spotted woodpecker (*Dendrocopos major*)](https://en.wikipedia.org/wiki/Great_spotted_woodpecker)
* [Coal tit (*Periparus ater*)](https://en.wikipedia.org/wiki/Coal_tit)
* [Common redstart (*Phoenicurus phoenicurus*)](https://en.wikipedia.org/wiki/Common_redstart)
* [Common chiffchaff (*Phylloscopus collybita*)](https://en.wikipedia.org/wiki/Common_chiffchaff)
* [Tree pipit (*Anthus trivialis*)](https://en.wikipedia.org/wiki/Tree_pipit)
* [Siskin (*Spinus spinus*)](https://en.wikipedia.org/wiki/Eurasian_siskin)
* [Hawfinch (*Coccothraustes coccothraustes*)](https://en.wikipedia.org/wiki/Hawfinch)
* [Bullfinch (*Pyrrhula pyrrhula*)](https://en.wikipedia.org/wiki/Eurasian_bullfinch)

### Exploring the data

Let's start by looking at a quick summariy of the dataset in order to get a feel for its content and structure.

```{r,  eval=T}
# dimension of the data frame birds
dim(birds)
# first three rows of the data frame birds
head(birds, 3 )
```

In this dataset, each row (or, each *observation*) represents the count of a species in a given year, on a given survey route.  The first row, for example, says that there were 2 individuals of *Alauda arvensis* observed in route number 1 during the year 1997.  Here's a description of the columns:

* *locationID* gives the route number
* *year* gives the year of the observation
* *species* gives the species observed
* *individualCount* gives the number of individuals observed
* *decimalLongitude* and *decimalLatitude* give the geographic coordinates for each route (in WGS84).
* *habitat* gives the main habitat association for the species (i.e., farmland or forest)

We might be interested to first get an aggregated summary of the data. For instance, let us look for each species at the total number of observations, that is, summed over all routes and over all years.  To do this, we construct a `data.frame` to hold information at the species-level. We do this in two steps. First, we simplify the data frame `birds` so that it only contains two columns, `species` and `habitat`.

```{r,  eval=T, results='hide'}
species.data <- unique(birds[ , c("species", "habitat")])
species.data
```

Notice, that this data frame lists first all farmland species and then all forest species. Second, to this simpliefied data frame, we want to add the total number of observations for each species. We use the `tapply` function to sum the counts of each species.

```{r,  eval=T, results='hide'}
# Sum counts for each species (over all years)
total.counts <- tapply(birds$individualCount, birds$species, sum)
total.counts
```

***Warning!***
Notice that the `total.counts` vector is in alphabetical order (this is the default behavior for `tapply` output). But the `species.data` data frame is sorted by habitat affiliation.  Thus, before we can add to total counts to the new data frame `species.data`, we have to change to order of the entries in `total.counts` to match the order of the species in `species.data`. We can do this using the `match` function.

```{r,  eval=T}
# We add a new column to the data frame `species.data` containing the total counts in the correct order
species.data$total.count <- total.counts[match(species.data$species, names(total.counts))]
species.data
```

Convince yourself that this produces the desired result. Let us visualize the total counts in a bar plot where the species are order by habitat affiliation.

```{r,  eval=T}
# Make a vector of colors for plotting
species.data$col <- ifelse(species.data$habitat %in% "Farms", "darkorange3", "forestgreen")

# Set figure margins for plotting
par(mar=c(13,4,1,1))

# Create a barplot of total counts per species
barplot(species.data$total.count/1000, 
        names.arg=species.data$species,
        las=2, col=species.data$col, 
        ylab="Total observations (x 1000)",
        cex.names=0.5)

# Add a legend to the plot
legend("topleft", legend=c("Farmland","Forest"), 
       pch=22, pt.bg=c("darkorange3", "forestgreen"),
       bty="n", pt.cex=2)
```

<span style="color: purple;">
**Question 1**: Please write down two conclusions you can make from this figure.
</span>

<!---
Next, we plot for each species the change in counts through time. To do so, we will use a programming tool called a `for loop`. *For loops* are iterative operations that loop over a program by using an index. If you are unfamiliar with *for loops* you can find a brief introduction [here](https://clausrueffler.github.io/CodingWithR/Scripting.html). It is probably easiest to digest with a simple example.

In the code below, the first line starts the `for loop` and says to use the index "year" for values 2010 through 2015. Everything inside the curly brackets `{ }` will be executed as the index, "year", iterates through the specified range of values. In this example, we ask R to print a phrase about which year it is. A pause of 0.5 seconds is included so that you can watch the iteration happening.

```{r,  eval=F}
for (year in 2010:2015){
  print(paste("The year is", year))
  Sys.sleep(0.5)
  }
```
That was a trivial example but hopefully you get the general idea. Let us try using a more complex *for loop* to plot the temporal trends for each bird species in our dataset.
--->

### Time series plots

Next, we plot for each species the change in counts through time. In order to do this we start by creating a new data frame in which we store the for each species and for each year the total count summed over all routes. We can do this using the `tapply`-function but now with respect to two factors, namely, species and year.

```{r eval = T, results = 'hide'}
count.per.year <- tapply(birds$individualCount, list(birds$year, birds$species), sum)

# Convice yourself that this gives the desired result.
head(count.per.year, 3)
```

Note, that again these counts are arranged alphabetically with respect to species names. We can use the same trick as above to rearrange the order so as to match the original order we use in the data frame `species.data`.

```{r eval=F}
count.per.year <- count.per.year[ , c(match(species.data$species, names(total.counts)))]

# Let us inspect whether the data look as intended.
head(count.per.year, 3)
```

Now we are ready to plot for each species the time series. To do so, we use a programming tool called a `for loop`. *For loops* are iterative operations that loop over a program by using an index. If you are unfamiliar with *for loops* you can find a brief introduction [here](https://clausrueffler.github.io/CodingWithR/Scripting.html).

```{r}
# First set the layout and margins of the figure
par(mfcol=c(5,4), mar=c(2,4,1.5,1), oma=c(0,4,0,4))

for (i in 1:ncol(count.per.year)){# this for-loop interatively produces the plot for the different species by cycling through the columns of the data frame `count.per.year`.
  plot(1996:2018, count.per.year[ , i], 
       type="l", ylab="Total Obs.", xlab=NA,
       col=ifelse(i %in% 1:10, "darkorange3", "forestgreen"), 
       lwd=3)
# Last, add a title at the top of the panel with the name of the species
  mtext(unique(birds$species)[i], 3, 0, cex=0.5, font=2)
}
```

<!---
```{r,  eval=T}
# Bob's original code
# First set the layout and margins of the figure
par(mfcol=c(5,4), mar=c(2,4,1.5,1), oma=c(0,4,0,4))

# Now we will use a `for loop` to cycle through each species.
# The first line says to use the index "i" for 1 through the 
# total number of bird species (20 in our case).
for (i in 1:length(unique(birds$species))){
# Next we subset the data corresponding to the i-th species
  focal_data <- birds[birds$species %in% unique(birds$species)[i],]
# Next we compute the counts per year for the i-th species
  trend <- tapply(focal_data$individualCount, focal_data$year, sum)  
# Now we plot the trend through time for the i-th species
  plot(names(trend), trend,
       type="l", ylab="Total Obs.", xlab=NA,
       col=ifelse("Farms" %in% focal_data$habitat, "darkorange3", "forestgreen"), 
       lwd=3)
# Last, add a title at the top of the panel with the name of the species
  mtext(unique(birds$species)[i], 3, 0, cex=0.5, font=2)
}
```
--->

Ok...what's going on? Are you surprised to see that the trends for all species appear to be increasing?  Are birds really taking over Sweden?!?  No, unfortunately not...

The general increase in total observations per species is at least partly due to the fact that the Swedish Bird Inventory added new survey routes during the first years so that the number of routes sampled per year is increasing. This is problematic for making valid comparisons between years because we are obviously likely to count more birds when we survey more routes (*think back to our lecture on* ***rarefaction***).

Now let us try and correct for the different number of routes sampled per year by computing the number of counts per species per year *per survey route*.

```{r, eval=F}
# Compute the number of routes recorded for each year
nroutes <- colSums(table(birds$locationID, birds$year) > 0)

# Let's redo the figure by dividing the number of observations 
# per species per year by the number of routes surveyed each year.
# First set the layout and margins of the figure.
par(mfcol=c(5,4), mar=c(2,4,1.5,1), oma=c(0,4,0,4))

for (i in 1:ncol(count.per.year)){
  plot(1996:2018, count.per.year[ , i]/nroutes, 
       type="l", ylab="Total Obs.", xlab=NA,
       col=ifelse(i %in% 1:10, "darkorange3", "forestgreen"), 
       lwd=3)
# Last, add a title at the top of the panel with the name of the species
  mtext(unique(birds$species)[i], 3, 0, cex=0.5, font=2)
}
```

<!---
```{r, eval=T}
# Bob's original code
# Compute the number of routes recorded for each year
nroutes <- colSums(table(birds$locationID, birds$year) > 0)

# Let's redo the figure by dividing the number of observations 
# per species per year by the number of routes surveyed each year
par(mfcol=c(5,4), mar=c(2,4,1.5,1), oma=c(0,4,0,4))

for (i in 1:length(unique(birds$species))){
  focal_data <- birds[birds$species %in% unique(birds$species)[i],]
  # Plot a panel for each species (note that we now divide by 'nroutes')
  plot(sort(unique(focal_data$year)),
  tapply(focal_data$individualCount, focal_data$year, sum)/nroutes, 
       type="l", ylab="Obs. per route", xlab=NA, lwd=3,
       col=ifelse("Farms" %in% focal_data$habitat, "darkorange3", "forestgreen"))
  mtext(unique(birds$species)[i], 3, 0, cex=0.5, font=2)
}
```
--->

<span style="color: purple;">
**Question 2**: After correcting for number of routes sampled per year, which species appears to have the most obvious increase in abundance over the survey period?  Which species appears to have the most obvious decrease in abundance over the survey period? Can you detect a general trend among farmland and forest species?
</span>

For times series like these it is common to report the data as an index relative to the count of the first year. Given the code that we have, it is easy to implement this change. All that we have to do is to divide the average number of individuals per route by the result for the first year. Let us have a look at the time series for the Skylark *Alauda arvensis*.

```{r eval=F}
# Time series for the average number of Skylarks per standard route for each year.
count.per.year[ , 1]/nroutes
```

The same timeseries but with all data divided by the result from the first year. Note. that now the count for the first year equals 1.

```{r eval=F}
(count.per.year[ , 1]/nroutes)/(count.per.year[1 , 1]/nroutes[1])
```

With this small modification. our time series plots can be computed as follows.

```{r, eval=F}
par(mfcol=c(5,4), mar=c(2,4,1.5,1), oma=c(0,4,0,4))

for (i in 1:ncol(count.per.year)){
  plot(1996:2018, (count.per.year[ , i]/nroutes)/(count.per.year[1 , i]/nroutes[1]),
       # for better comparision with the plots from Svensk Fågeltaxering we change the y-axis so that it starts at zero and is exactly as long as needed 
       ylim = c(0, max((count.per.year[ , i]/nroutes)/(count.per.year[1 , i]/nroutes[1])+0.3)), yaxs="i",
       type="l", ylab="Total Obs.", xlab=NA,
       col=ifelse(i %in% 1:10, "darkorange3", "forestgreen"), 
       lwd=3)
  mtext(unique(birds$species)[i], 3, 0, cex=0.5, font=2)
}
```

<span style="color: purple;">
**Question 3**: 
Let us compare our plots with those produced by Svensk Fågeltaxering. You can find the official time series plot from Svensk Fågeltaxering [here](https://www.fageltaxering.lu.se/resultat/trender/trender-standardrutterna).
What do you conclude?
</span>

